---
title: "MX"
format: pdf
editor: visual
---

# Masseeksperimentet (+MFD)

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = FALSE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  fig.height = 8,
  fig.width = 8
)

# use renv from the lock file
if (!require("renv")) {
  install.packages("renv")
}
renv::load()

library("data.table")
library("readxl")
library("dplyr")
library("tidyr")
library("tibble")
library("ampvis2")
library("Nonpareil")
library("patchwork")
library("foreach")
library("doFuture")
library("tidyverse")
library("RColorBrewer")

shinyapp_dir <- "app"
# dir for plots for the app (relative to this file, NOT the app dir)
shinyapp_plots_dir <- "plots"
# delete these files to regenerate data
# fx: unlink(MX_ampvis2_obj_out)
MX_ampvis2_obj_out <- "data/MX/mx.rds"
MFD_ampvis2_obj_out <- "data/MFD/mfd.rds"

# Lower this if you don't have enough memory, or set the future::plan below to sequential
# See https://www.r-bloggers.com/2023/06/dofuture-a-better-foreach-parallelization-operator-than-dopar/
maxthreads <- as.integer(system("nproc", intern = TRUE))
setDTthreads(maxthreads)
doFuture::registerDoFuture()
plan(multisession, workers = maxthreads)
```

## MX metadata

```{r loaddata}
if (!file.exists(MX_ampvis2_obj_out)) {
  # let's make a decision about naming the sample ID column consistently everywhere!!
  sampleid_colname <- "SampleID"

  ### sequencing metadata with libIDs etc
  seq_metadata <- fread("data/MX/metadata/data_overeview(Ark1).csv")
  colnames(seq_metadata)[1] <- sampleid_colname
  seq_metadata[[1]] <- gsub("[^0-9]+", "", seq_metadata[[1]])
  seq_metadata <- seq_metadata[!is.na(eval(parse(text = sampleid_colname))) & !is.na(LibID) & !duplicated(LibID)]

  ### nonpareil, merge with npo filenames, obtain numreads per fastq file, calc nonpareil "diversity"
  npofiles <- data.table(
    npofile = list.files("data/MX/nonpareil/output", pattern = "*.npo", recursive = TRUE, full.names = T)
  )
  npofiles[, LibID := gsub("_R[0-9]+.*$", "", tools::file_path_sans_ext(basename(npofile)))]
  getreadsfromnpofile <- function(npofile) {
    npofile <- readLines(npofile, n = 10)
    numreads <- as.integer(gsub("[^0-9]+", "", npofile[grepl("^# @R", npofile)]))
    return(numreads)
  }
  npofiles[, numreads := getreadsfromnpofile(npofile), by = npofile]
  npofiles[, nonpareil_diversity := tryCatch(
    {
      Nonpareil.curve(npofile, plot = FALSE)@diversity
    },
    error = function(e) {
      return(NA)
    }
  ), by = npofile]
  npofiles[nonpareil_diversity == 0, nonpareil_diversity := NA]
  metadata <- npofiles[seq_metadata, on = "LibID"]

  ### original/official metadata
  metadata_off <- read_excel("data/MX/metadata/MasterFile_MX2025_20260320.xlsx")
  setDT(metadata_off)
  colnames(metadata_off)[1] <- sampleid_colname
  metadata_off[[sampleid_colname]] <- gsub("[^0-9]+", "", metadata_off[[sampleid_colname]])
  metadata <- metadata_off[metadata, on = sampleid_colname]
  metadata <- metadata[, .SD, by = "LibID"] # move LibID column to first column

  # need habitat type to do any analysis
  metadata <- metadata[!is.na(HabitatType)]
}
```

## MX abundance data

```{r}
if (!file.exists(MX_ampvis2_obj_out)) {
  ### ABUNDANCE DATA
  # sylph output
  taxrelabund <- fread("data/MX/MASSE160326_quant/MASSE160326_tax_relative_abundance.tsv.zip")
  colnames(taxrelabund)[colnames(taxrelabund) %in% "clade_name"] <- "OTU"
  taxrelabund_genus <- taxrelabund[grepl("g[_]+[^|]+$", OTU)]

  # for some reason there are taxa with multiple parents
  Genera <- gsub(".*g__", "", taxrelabund_genus[["OTU"]])
  polyphyletics <- taxrelabund_genus[["OTU"]][duplicated(Genera) | duplicated(Genera, fromLast = TRUE)]
  taxrelabund_genus <- taxrelabund_genus[!taxrelabund_genus[["OTU"]] %chin% polyphyletics]

  # generate taxonomy table from clade_name
  tax <- separate_wider_delim(
    taxrelabund_genus[, 1, drop = FALSE],
    "OTU",
    delim = "|",
    names = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus"),
    cols_remove = FALSE
  )
  tax[["Species"]] <- tax[["Genus"]] -> tax[["OTU"]]
  # amp_rename also removes g__ prefixes from OTU col (it should NOT, TODO!) which causes trouble
  tax[["OTU"]] <- gsub("^[dkpcofgst]__", "", tax[["OTU"]])

  # extract abundance table and sample IDs from filenames
  abund <- taxrelabund_genus
  abund[["OTU"]] <- gsub(".*[dkpcofgst]__", "", abund[["OTU"]])
  colnames(abund) <- gsub("_R1.fastq.gz$", "", basename(colnames(abund)))

  # load as ampvis2 format
  MX <- amp_load(
    otutable = abund,
    taxonomy = tax,
    metadata = metadata
  )
  saveRDS(MX, MX_ampvis2_obj_out)
}
```

## MFD metadata

```{r}
metadata_filepath <- "data/MFD/metadata.xlsx" # delete file to update
if (!file.exists(metadata_filepath)) {
  download.file(
    "https://github.com/cmc-aau/mfd_metadata/raw/refs/heads/main/analysis/releases/latest_mfd_db.xlsx",
    destfile = metadata_filepath,
    mode = "wb"
  )
}
metadata <- read_xlsx(metadata_filepath)
```

## MFD abundance data

```{r}
if (!file.exists(MFD_ampvis2_obj_out)) {
  ### ABUNDANCE DATA
  # sylph output
  taxrelabund <- fread(
    "data/MFD/MFD_GTDB_R226_quant/formatted/MFD_GTDB_R226_tax_relative_abundance_data.tsv.zip"
  )
  sampleid_col <- colnames(taxrelabund)[1]
  # extract genus-level abundances and tax
  taxrelabund_genus <- taxrelabund[
    ,
    c(1, grep("g[_]+[^|]+$", colnames(taxrelabund))),
    with = FALSE
  ]
  # for some reason there are taxa with multiple parents
  Genera <- gsub(".*g__", "", colnames(taxrelabund_genus))
  polyphyletics <- colnames(taxrelabund_genus)[duplicated(Genera) | duplicated(Genera, fromLast = TRUE)]
  taxrelabund_genus <- taxrelabund_genus[, !colnames(taxrelabund_genus) %chin% polyphyletics, with = FALSE]

  # generate taxonomy table from "OTU" (=Genus) names
  tax <- separate_wider_delim(
    data.table(OTU = colnames(taxrelabund_genus)[-1]),
    "OTU",
    delim = "|",
    names = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus"),
    cols_remove = FALSE
  )
  # use Genus as OTU ID's, not the full lineage
  tax[["Species"]] <- tax[["Genus"]] -> tax[["OTU"]]
  # amp_rename also removes g__ prefixes from OTU col (it should NOT, TODO!) which causes trouble
  tax[["OTU"]] <- gsub("^[dkpcofgst]__", "", tax[["OTU"]])

  # cast abundance table
  taxrelabund_genus %>%
    melt(
      id.vars = sampleid_col,
      variable.name = "OTU"
    ) %>%
    dcast(OTU ~ eval(parse(text = sampleid_col))) %>%
    {
      .[["OTU"]] <- gsub(".*g__", "", .[["OTU"]])
      .
    } -> abund

  MFD <- amp_load(
    otutable = abund,
    taxonomy = tax,
    metadata = metadata
  )
  saveRDS(MFD, MFD_ampvis2_obj_out)
}
```

## Load ampvis2 objects

```{r}
if (!exists("MX", envir = .GlobalEnv)) {
  MX <- readRDS(MX_ampvis2_obj_out)
}
if (!exists("MFD", envir = .GlobalEnv)) {
  MFD <- readRDS(MFD_ampvis2_obj_out)
}
```

# From Mads

## Filter samples

First we only take 1 sample pr. 1km2 and also prioritize the 10km2 representative samples.

```{r}
cell1km <- MFD$metadata %>%
  group_by(mfd_sampletype, mfd_hab1, mfd_hab2, mfd_hab3, cell.1km) %>%
  arrange(desc(cell.10km_representative)) %>%
  mutate(count = n()) %>%
  filter(count == 1)
```

Then we subset to relevant samples and remove strange samples.

```{r}
mfdf <- amp_filter_samples(
  data = MFD, after_total_reads > 10000000, after_total_reads < 100000000,
  fieldsample_barcode %in% cell1km$fieldsample_barcode,
  mfd_areatype != "Subterranean",
  !(mfd_hab1 %in% c(
    "Biogas", "Drinking water", "Landfill", "Roadside",
    "Rocky habitats and caves", NA, "Saltwater",
    "Wastewater", "Freshwater", "Coastal", "Dunes",
    "Bogs, mires and fens"
  )),
  !(mfd_hab2 %in% c(
    NA, "Mire (non-habitat type)", "Asparagales",
    "Superasterids", "Other", "Influent", "Asterids",
    "Fabids", "Flowing saltwater", "Forest (non-habitattype)",
    "Natural grasslands", "Inland dunes",
    "Wet thicket  (non-habitat type)", "Fallow", "Malvids",
    "Mixed crops", "Scrub"
  )),
  !(mfd_hab3 %in% c(
    "Beech forests Acidophilous with Ilex",
    "Beech forests Limestone",
    "Natural meadow (6410 subtype)",
    "Oak-hornbeam forests Galio-Carpinetum", "Bog woodland",
    "Wheat, Winter, Green Grain", "Barley, Spring,
                                             Green grain", "Maize", ""
  )),
  !(is.na(mfd_hab3) & mfd_hab1 == "Forests"),
  !(is.na(mfd_hab3) & mfd_hab2 == "Poales, Cereal"),
  !(is.na(mfd_hab3) & mfd_hab2 == "Poales, grass"),
  !(is.na(mfd_hab3) & mfd_hab1 == "Grassland formations"),
  !(fieldsample_barcode %in% c(
    "MFD04182", "MFD00149", "MFD04817", "MFD05018",
    "MFD00357", "MFD05367", "MFD06769", "MFD05703",
    "MFD00282", "MFD00415", "MFD02404"
  ))
)
```

We rename the mfd2 catagories to something the students might understand.

```{r}
namechange <- mfdf$metadata %>%
  mutate(
    mfd_hab2 = ifelse(mfd_hab3 != "Alluvial woodland" & mfd_hab2 == "Temperate forests",
      "Blandet skov", mfd_hab2
    ),
    mfd_hab2 = ifelse(mfd_hab3 == "Alluvial woodland" & mfd_hab2 == "Temperate forests",
      "Skove ved vandløb og søer", mfd_hab2
    ),
    mfd_hab2 = ifelse(mfd_hab2 == "Poales, Cereal", "Kornmarker", mfd_hab2),
    mfd_hab2 = ifelse(mfd_hab2 == "Poales, grass", "Græsmarker", mfd_hab2),
    mfd_hab2 = ifelse(mfd_hab2 == "Parks", "Parker", mfd_hab2),
    mfd_hab2 = ifelse(mfd_hab2 == "Temperate heath", "Tør hede", mfd_hab2),
    mfd_hab2 = ifelse(mfd_hab2 == "Semi-natural dry grasslands", "Kalk- og sure overdrev", mfd_hab2),
    mfd_hab2 = ifelse(mfd_hab2 == "Semi-natural tall-herb humid meadows", "Bevoksning langs vandløb", mfd_hab2)
  )

mfdf$metadata <- namechange
```

We check the number of samples in each category.

```{r}
mfdf$metadata %>%
  group_by(mfd_hab1, mfd_hab2) %>%
  count() %>%
  print(n = 200)
```

## Load mx data

```{r}
mxf <- amp_filter_samples(data = MX, Reseq != "x", HabitatType != "kyst", !is.na(Region))
```

## Merge

```{r}
mm <- amp_merge_ampvis2(mxf, mfdf, by_refseq = FALSE)
```

## Filter species

```{r}
species <- mm$abund %>%
  mutate(genus = rownames(mm$abund)) %>%
  pivot_longer(cols = 1:(ncol(mm$abund) - 1)) %>%
  filter(value > 0.01) %>%
  group_by(genus) %>%
  summarise(
    median = median(value),
    mean = mean(value),
    max = max(value),
    count = n()
  ) %>%
  arrange(desc(count)) %>%
  filter(count > 20) %>%
  filter(max > 0.1)

mms <- amp_filter_taxa(mm, tax_vector = species$genus)
```

## Add metadata

```{r}
mms2 <- mms
mms2$metadata <- mms$metadata %>%
  mutate(
    kategori = ifelse(is.na(HabitatType), mfd_hab2, "Masse eksperimentet"),
    mxdata = ifelse(is.na(HabitatType), "MicroFlora Danica", "MX"),
    katskole = ifelse(is.na(HabitatType), mfd_hab2, LibID)
  )

# write out metadata for shiny app, reduce to an absolute minimum to reduce app.json size
fwrite(mms2$metadata[, c("LibID", "SampleID", "PointLongitude", "PointLatitude")], file.path(shinyapp_dir, "metadata.csv"))
```

## Heatmap - overordnet

```{r}
p_heatmap_master <- amp_heatmap(mms2,
  tax_aggregate = c("Phylum", "Genus"),
  tax_show = 25,
  group_by = "kategori",
  facet_by = "mxdata",
  plot_colorscale = "sqrt",
  plot_values_size = 4,
  normalise = FALSE
) +
  theme(
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    strip.text.x = element_text(size = 8),
    legend.position = "none"
  )

# ggsave(filename = "heatmap.png", width = 14, height = 14, units = "cm", dpi = 300)
```

# Ordinering - overordnet

```{r}
# master ordination on which to add/highlight individual sample points
p_ord_master <- amp_ordinate(
  mms2,
  sample_color_by = "kategori",
  filter_species = 0,
  transform = "sqrt",
  constrain = "kategori",
  type = "RDA"
)
p_ord_master$layers[[1]] <- NULL
```

# Generate plots per sample for shiny app

```{r}
# Sample IDs 1087 and 1514 are duplicated!
# use a dynamic ID col, could also be SampleID or any other var
idcol <- "LibID"
ids <- mms2$metadata[mms2$metadata$mxdata == "MX", idcol]
heatmaps <- foreach(
  id = ids, # just subset here for developing
  .combine = c,
  .inorder = TRUE,
  # export only a bare minimum of objects and pkgs to all the forked processes to minimize memory usage
  .options.future = list(
    globals = c("mms2", "idcol", "shinyapp_dir", "shinyapp_plots_dir", "p_ord_master"),
    packages = c("ampvis2", "patchwork", "RColorBrewer")
  )
) %dofuture% {
  d <- suppressMessages(amp_filter_samples(mms2, eval(parse(text = idcol)) == id))
  
  # heatmap, sample
  p_heatmap_sample <- amp_heatmap(
    d,
    tax_aggregate = c("Phylum", "Genus"),
    tax_show = 25,
    group_by = "katskole",
    facet_by = idcol,
    plot_colorscale = "sqrt",
    plot_values_size = 4,
    normalise = FALSE
  ) +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 12),
    strip.text.x = element_text(size = 10),
    legend.position = "none"
  )
  # heatmap, MX+MFD subset to sample taxa
  p_heatmap_master_subset <- amp_heatmap(
    mms2,
    tax_aggregate = c("Phylum", "Genus"),
    tax_show = levels(droplevels(p_heatmap_sample$data$Display)),
    order_y_by = levels(droplevels(p_heatmap_sample$data$Display)),
    group_by = "kategori",
    facet_by = "mxdata",
    plot_colorscale = "sqrt",
    plot_values_size = 4,
    normalise = FALSE
  ) +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.text.x = element_text(size = 14),
      strip.text.x = element_text(size = 10),
      legend.position = "none"
    )
  p_heatmap <- p_heatmap_sample +
    p_heatmap_master_subset +
    plot_layout(
      nrow = 1,
      widths = c(1.5, 8.5)
    )

  # ordination
  p_ord_sample <- p_ord_master +
    stat_ellipse(
      data = p_ord_master$data %>% subset(mxdata == "MicroFlora Danica"),
      level = 0.5,
      size = 1,
      geom = "polygon",
      aes(
        fill = kategori,
        color = kategori
      ),
      alpha = 0.7
    ) +
    geom_point(
      data = p_ord_master$data %>% subset(mxdata == "MX"),
      aes(shape = HabitatType), color = "black", alpha = 0.2, size = 3
    ) +
    scale_color_manual(values = brewer.pal(9, "Set1"), name = "MicroFlora Danica") +
    scale_fill_manual(values = brewer.pal(9, "Set1"), name = "MicroFlora Danica") +
    scale_shape_manual(name = "Masse Eksperimentet", values = c(16, 17, 15, 18)) +
    geom_point(
      data = p_ord_master$data %>% subset(eval(parse(text = idcol)) == id),
      aes(shape = HabitatType), size = 4, color = "red", alpha = 1
    ) +
    theme(
      legend.text = element_text(size = 8),
      legend.title = element_text(size = 10)
  )
  
  ggsave(
    file.path(shinyapp_plots_dir, paste0(mms2$metadata[mms2$metadata[[idcol]] %in% id, idcol], "_heatmap.png")),
    plot = p_heatmap,
    create.dir = TRUE,
    height = 8,
    width = 10,
    dpi = 200
  )

  ggsave(
    file.path(shinyapp_plots_dir, paste0(mms2$metadata[mms2$metadata[[idcol]] %in% id, idcol], "_ordination.png")),
    plot = p_ord_sample,
    create.dir = TRUE,
    height = 8,
    width = 8,
    dpi = 200
  )
  invisible(plot)
}
```

# Misc

```{r}
# heatmap per habitat
amp_heatmap(
  MX,
  tax_aggregate = "Genus",
  tax_show = 20,
  group_by = "HabitatType",
  normalise = FALSE
)
```

```{r}
# ordination
amp_ordinate(
  MX,
  type = "nmds",
  distmeasure = "bray",
  sample_color_by = "HabitatType",
  filter_species = 0,
  plot_species = TRUE,
  species_nlabels = 30
)
```

```{r}
# nonpareil
ggplot(
  MX$metadata,
  aes(
    y = nonpareil_diversity,
    group = HabitatType,
    color = HabitatType
  )
) +
  geom_boxplot()
```
